Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-01-12 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Correction
changes on progress

\begin{abstract} 
  The aim of this paper is to model flow instabilities \ie viscous fingering using a novel method for modeling multi-phase flows in heterogeneous porous media. Starting from modelling the Hele-Shaw experimental apparatus, which is mathematically analogous to the two-dimensional flow in porous media problem. The development of finger structures is investigated by taking into account different conditions such as, different domains under uniform or different permeabilities regions, mobility ratios, mesh types. A novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion. It was found that mobility ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an ustructured mesh allow the user to better describe and visualise the phenomenon.   \red{KOSTAS: Abstracts are divided into 4 parts: (1) Motivation (1-2 sentences); (2) Main aim of the manuscript highlighting the novelty of the work (you must not use the word novelty ... this should be implicit); (3) summary of the main technologies/methods and; (4) summary of the main findings.}

\end{abstract}

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
 Multi-fluid flows \sep Porous media \sep Viscous Instabilities \sep Mobility Ratio \sep Preferential flowpath \sep Finite Element Methods \sep Capillary number \sep Unstructured domain \sep Mesh adaptivity. \red{KOSTAS: choose 3-5 key-words that better represent the work done.}
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, different permeability characteristics in different zones) of geological formations. In this application, volumetric sweep efficiency \red{(KOSTAS: MAY BE YOU SHOULD ADD ONE SENTENCE DEFINING/EXPLAINING ``SWEEP EFFICIENCY'' AND ``BREAKTHROUGH'' ... AND HOW THEY ARE RELATED.)} may be reduced and early breakthrough may occur \citep{riaz_2004, tavassoli_2015}, leading to low oil production. During immiscible CO$_{2}$ flooding $\left(\text{\eg in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, density of supercritical CO$_{2}$ is usually lower than hydrocarbons, thus viscous fingering and/or channelling may occur.

\medskip
Flow instabilities may be due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain. %\red{For example, during water flooding, efficiency is related with capillary number. The higher the capillary number the higher the recovery of hydrocarbons (LINK THESE SENTENCES ! HOW THIS NEW SENTENCE IS RELATED TO THE PREVIOUS -- 'Flow instabilities ... of the domain' ... AND HOW IS IT RELATED WITH THE FOLLOWING SENTENCES?)}. 
When the driving force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and start forming fingers. \red{(KOSTAS: YOU STILL NEED TO REWRITE THIS PARAGRAPH AND LINK BOTH SENTENCES. THE SECOND SENTENCE (``When the driving'') IS A GOOD START FOR THE PARAGRAPH, ALTHOUGH YOU SHOULD DETERMINE WHAT IS THE ''DRIVING FORCE''  THAT YOU REFERRED TO.)}

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model capability to capture viscous crossflow in immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial discontinuities of permeability distribution will create a rough pressure field, \ie the pressure gradient is discontinuous across the interface between different permeability zones thus the flow will be lead to the high permeability region. \red{(PS: I AM NOT SURE WHICH INFO YOU WANT TO CONVEY IN THIS SENTENCE (``The spatial discontinuities ... permeability region'') ... ACTUALLY THIS SENTENCE IS NOT CORRECT AS (IN THIS CASE) THERE IS NO DISCONTINUITY IN EITHER PERMEABILITY OR IN PRESSURE, JUST A LARGE K AND P DISTRIBUTION ... IN ADDITION ...\underline{IS THERE ANY COMMENT WRT FIG 4b-c AT VR=1?})}

>>>>

\begin{abstract} 
  The aim of this paper is to model flow instabilities \ie viscous fingering using a novel method for modeling multi-phase flows in heterogeneous porous media. Starting from modelling the Hele-Shaw experimental apparatus, which is mathematically analogous to the two-dimensional flow in porous media problem. The development of finger structures is investigated by taking into account different conditions such as, different domains under uniform or different permeabilities regions, mobility ratios, mesh types. A novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion. It was found that mobility ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an ustructured mesh allow the user to better describe and visualise the phenomenon.   \red{KOSTAS: Abstracts are divided into 4 parts: (1) Motivation (1-2 sentences); (2) Main aim of the manuscript highlighting the novelty of the work (you must not use the word novelty ... this should be implicit, 2-6 sentences); (3) summary of the main technologies/methods (2-5 sentences) and; (4) summary of the main findings (1-3 sentences).}     
\end{abstract}

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
 Multi-fluid flows \sep Porous media \sep Viscous Instabilities \sep Mobility Ratio.
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Brief Summary of Viscous Instabilities}\label{section:ViscousInstabilities}
The study of viscous flow instabilities (\ie fingering) is particularly important in oil exploration due to heterogeneities (\ie natural fractures, different permeability characteristics in different zones) of geological formations that may affect the oil production. Important factors during flooding operations are the volumetric sweep efficiency, or the ration of volumes between the recovered and the injected fluids as well as the breakthrough of injected fluid. A major problem associated with water-flooding projects is the early water-breakthrough, the main causes of which are high-permeability layers and unfavorable mobility ratios. Water breakthrough is the main determinant of the productive life of the reservoir \citep{riaz_2004, tavassoli_2015}. During immiscible CO$_{2}$ flooding $\left(\text{\eg in CO}_{2}\text{ enhanced oil recovery, CO}_{2}\text{-EOR, processes}\right)$, density of supercritical CO$_{2}$ is usually lower than hydrocarbons, thus viscous fingering and/or channelling may occur.

\medskip
The displacement efficiency depends upon the ratio of the viscous to capillary forces or capillary number. 
%Flow instabilities will occur due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain (layering). 
When viscous force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and start forming fingers.

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model capability to capture viscous crossflow in immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial discontinuities of permeability distribution will create a rough pressure field, \ie the pressure gradient is discontinuous across the interface between different permeability zones thus the flow will be lead to the high permeability region. \red{{IS THERE ANY COMMENT WRT FIG 4b-c AT VR=1?}}
